A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-05-11 tex/REBUTTAL.txt

Reviewer 1:

Regarding the evaluation, analysis of raw data http://lara.epfl.ch/~reynolds/CADE2015-cdt/ shows that the overhead of using the more expensive rules (Acyclic and Unique) is negligible for this set. The 1059 benchmarks solved by both CVC4 (full datatypes) and CVC4 (datatypes without Acyclic) are solved collectively in 48.7 and 48.4 seconds respectively, and the 1082 benchmarks solved by both CVC4 (full codatatypes) and CVC4 (codatatypes without Unique) are solved in 54.0 and 54.1 seconds respectively. Notice that these rules may increase runtime due to the time spent on computing when they are applicable, but may also decrease runtime due to the solver's ability to find a smaller proof. We will comment on this point in the evaluation.

The statement "the stronger decision procedures almost completely subsume the weaker ones in practice" references the fact that in only 1 case (out of 4170) did a configuration with a weaker decision procedure solve a benchmark that was unsolved by a stronger one. Such outliers are common for benchmarks containing universally quantified formulas, since as we mention in Section 4, the SMT solver is generally incomplete when such formulas are present. We will make this statement explicit.

Reviewer 2 and 3:

We should clarify the relationship between this paper and Barrett et al [2], which discussed an implementation of the datatypes theory solver in CVC3. Regarding implementation, CVC4 was a from-scratch rewrite of CVC3, and does not share any code used in [2]. These two implementations share the same ideas, but differ significantly in practice, even when restricted to datatypes only. For instance, the implementation in CVC4 is built on top of a general-purpose data structure for efficient congruence closure that represents equivalence classes as circular lists, whereas the one in CVC3 uses a union-find data structure. Some of these differences are implicitly reflected in the calculus, in that [2] describes the use of oriented equations, whereas this paper does not. The datatypes theory solver in CVC4 is (as of yet) unpublished work. We will elaborate on this relationship in the paper.

Reviewer 2:

While the addition of codatatypes adds only two rules to the calculus, in our opinion, in terms of metatheory, the difference is significant. Most notably, the proof of completeness in this paper is not a straightforward lifting of the proof completeness in [2]. Since the proof from Section 4 of [2] does not include codatatypes, the proof of completeness in this paper required both a different formalization of values (as closed normal mu-terms), and new reasoning to recognize the set of (possibly cyclic) values V^x_I that may lead us to non-models when constructing A*.

We will add more details regarding the role of A and A* (notice they are required not only to argue for correctness, but also for the SMT solver to be model-producing). We will also comment more on how they are implemented, which we will reference earlier in the paper.

### 2015-05-11 tex/body.tex
Address reviewer comments.

\ourparagraph{Phase 2: Applying Acyclicity and Uniqueness \beforefig(\figurename~\ref{fig:ab-rules}).}
%For presentation of the rules in this phase, we rely on a representation of terms in the $\mu$-notation.
The premises of the rules that make up the second phase refer to a mapping $\Val$
that associates with each equivalence class a $\mu$-term as its representative.
%
Formally, $\mu$-\emph{terms} are defined recursively as being either a variable $x$
or an applied constructor
$\MU x.\: \const{C}( \vec t\, )$ for some $\const{C} \in \Ctr$ and
$\mu$-terms $\vec t$ of the expected types.
The variable~$x$ need not occur free in the $\mu$-binder's body, in which case
the binder can be omitted.

$\FV( t )$ denotes the set of free variables occurring in the $\mu$-term~$t$.
A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. It is \emph{cyclic} if %and only if --- ditto
it contains a bound variable. The $\alpha$\emph{-equiv\-alence} relation $t \aequiv u$
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables\begin{rep}---e.g.,
%$\MU x.\; \const{C}( x ) \aequiv \MU y.\; \const{C}( y )$ and
$\MU x.\; \const{D}( y, x ) \aequiv \MU z.\; \const{D}( y, z )$,
but
$\MU x.\; \const{C}( x ) \not\aequiv \MU x.\; \const{D}( y, x ) \not\aequiv \MU x.\; \const{D}( z, x )
\not\aequiv \MU y.\; \const{D}( y, x )$\end{rep}.
Two $\mu$-terms can denote the same value despite being $\alpha$-disequivalent---e.g.,
$\MU x.\; \const{S}( x ) \not\aequiv \MU y.\; \const{S}( \const{S}( y ) )$.
%In the \rn{Acyclic} rule,

@@

The mapping $\Val$ is computed as follows.
%
With each equivalence class $\ec{u\typ{\tau}}$, we associate a fresh variable $\Varec{u}\typ{\tau}$
%of the same type as $u$
%not occurring in $\tEc$,
and set $\Val\ec{u} := {\Varec{u}}$, that is to say there are initially no
constraints on the values for any equivalence class $\ec{u}$. The mapping $\Val$
is refined by applying the following unfolding rule exhaustively:\strut
\[
\hbox{\(
\inferrule{
  {\Varec{u}} \in \FV( \Val )
  \betweenantes
  \const{C}( t_1, \ldots, t_n ) \in \ec{u}
  \betweenantes
  \const{C} \in \Ctr
}{
  \Val := \Val [{\Varec{u}} \mapsto \MU {\Varec{u}}.\; \const{C}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} )]
}
\)
%\rn{Unfold}
}
\]
$\FV( \Val )$ denotes the set of free variables occurring in $\Val$'s range,
and $\Val[x \mapsto t]$ denotes the \emph{variable-capturing} substitution of $t$ for
$x$ in $\Val$'s range. It is easy to see that the height of terms produced as a
result of the unfolding is bounded by the number of equivalence classes of
$\Ec$, and thus the construction of $\Val$ will terminate.

@@

By Theorems~\ref{thm:t},~\ref{thm:rs}, and~\ref{thm:ss}, the
calculus is sound and complete for the universal theory of (co)datatypes. We can
rightly call it a decision procedure for that theory.

@@

An evaluation of the efficiency of the procedure is described by Barrett et
al.\ \cite{barrett-et-al-2007}. In particular, it is both theoretically and
empirically beneficial to delay applications of \rn{Split} as long as
possible. Similarly, \rn{Acyclic} and \rn{Unique} are fairly expensive because
they require traversing the equivalence classes, which is why they are part of
phase~2.

@@

The results are summarized in Table~\ref{tab:bench}. The decision procedure
makes a difference across all three benchmark suites. It accounts for an overall
success rate increase of over \OK{5\%}, which is quite significant. For Isabelle
users, the practical usefulness of interpreted (co)datatypes is roughly
comparable to that of arithmetic \cite{blanchette-et-al-2013-smt}, although this
obviously depends on the application. The raw evaluation data also
suggest that the theoretically stronger decision procedures almost completely
subsume the weaker ones in practice.

>>>>

\ourparagraph{Phase 2: Applying Acyclicity and Uniqueness \beforefig(\figurename~\ref{fig:ab-rules}).}
%For presentation of the rules in this phase, we rely on a representation of terms in the $\mu$-notation.
%The premises of the rules that make up the second phase refer to
We describe the rules in this phase in terms of a mapping $\Val$
that associates with each equivalence class a $\mu$-term as its representative.
\rem{Not explicitly constructed, add here?}

Formally, $\mu$-\emph{terms} are defined recursively as being either a variable $x$
or an applied constructor
$\MU x.\: \const{C}( \vec t\, )$ for some $\const{C} \in \Ctr$ and
$\mu$-terms $\vec t$ of the expected types.
The variable~$x$ need not occur free in the $\mu$-binder's body, in which case
the binder can be omitted.
$\FV( t )$ denotes the set of free variables occurring in the $\mu$-term~$t$.
A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. It is \emph{cyclic} if %and only if --- ditto
it contains a bound variable. The $\alpha$\emph{-equiv\-alence} relation $t \aequiv u$
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables\begin{rep}---e.g.,
%$\MU x.\; \const{C}( x ) \aequiv \MU y.\; \const{C}( y )$ and
$\MU x.\; \const{D}( y, x ) \aequiv \MU z.\; \const{D}( y, z )$,
but
$\MU x.\; \const{C}( x ) \not\aequiv \MU x.\; \const{D}( y, x ) \not\aequiv \MU x.\; \const{D}( z, x )
\not\aequiv \MU y.\; \const{D}( y, x )$\end{rep}.
Two $\mu$-terms can denote the same value despite being $\alpha$-disequivalent---e.g.,
$\MU x.\; \const{S}( x ) \not\aequiv \MU y.\; \const{S}( \const{S}( y ) )$.
%In the \rn{Acyclic} rule,

@@

The mapping $\Val$ is defined as follows.
%
With each equivalence class $\ec{u\typ{\tau}}$, we associate a fresh variable $\Varec{u}\typ{\tau}$
%of the same type as $u$
%not occurring in $\tEc$,
and set $\Val\ec{u} := {\Varec{u}}$, that is to say there are initially no
constraints on the values for any equivalence class $\ec{u}$. The mapping $\Val$
is refined by applying the following unfolding rule exhaustively:\strut
\[
\hbox{\(
\inferrule{
  {\Varec{u}} \in \FV( \Val )
  \betweenantes
  \const{C}( t_1, \ldots, t_n ) \in \ec{u}
  \betweenantes
  \const{C} \in \Ctr
}{
  \Val := \Val [{\Varec{u}} \mapsto \MU {\Varec{u}}.\; \const{C}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} )]
}
\)
%\rn{Unfold}
}
\]
$\FV( \Val )$ denotes the set of free variables occurring in $\Val$'s range,
and $\Val[x \mapsto t]$ denotes the \emph{variable-capturing} substitution of $t$ for
$x$ in $\Val$'s range. It is easy to see that the height of terms produced as a
result of the unfolding is bounded by the number of equivalence classes of
$\Ec$, and thus the construction of $\Val$ will terminate.

@@

By Theorems~\ref{thm:t},~\ref{thm:rs}, and~\ref{thm:ss}, the
calculus is sound and complete for the universal theory of (co)datatypes. We can
rightly call it a decision procedure for that theory.
We remark that the proof of solution soundness is constructive,
meaning that it provides a method for constructing a model for a saturated configuration, by means of the mapping $\ValC$.

@@

% AJR : this paragraph may have been what confused Reviewer #2. It does suggest that we are copying verbatim what was in CVC3 (which is what Reviewer #2 thought).
% Since we aren't saying anything new beyond what is said in Barrett et al (and honestly, we didn't critically evaluate the performance of the updated implementation), 
% maybe we should just remove this paragraph.

%An evaluation of the efficiency of the procedure is described by Barrett et
%al.\ \cite{barrett-et-al-2007}. In particular, it is both theoretically and
%empirically beneficial to delay applications of \rn{Split} as long as
%possible. Similarly, \rn{Acyclic} and \rn{Unique} are fairly expensive because
%they require traversing the equivalence classes, which is why they are part of
%phase~2.

@@

The results are summarized in Table~\ref{tab:bench}. The decision procedure
makes a difference across all three benchmark suites. It accounts for an overall
success rate increase of over \OK{5\%}, which is quite significant. For Isabelle
users, the practical usefulness of interpreted (co)datatypes is roughly
comparable to that of arithmetic \cite{blanchette-et-al-2013-smt}, although this
obviously depends on the application. The raw evaluation data also
suggest that the theoretically stronger decision procedures almost completely
subsume the weaker ones in practice.
In particular, we encountered only one benchmark over the entire set of 4170 benchmarks
that was solved by a configuration of CVC4 and unsolved by a configuration of CVC4 with more features enabled.
